From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-27 upi.tex
progress on sec 2

\jacques{Example 1: using the combinators from pi-dual, some an equational
derivation in type theory (which?), in equational-derivation style.  Point out
that reading ``down'' the justifications is a $\Pi$ program which is also a
constructive proof of the equivalence of the types.  As it is reversible, we
can also read the reverse program by reading ``up''.}

\newcommand{\byiso}[1]{{\leftrightarrow}{\langle} ~#1~ \rangle}

{\small
\[\def\arraystretch{1.2}\begin{array}{rcl}
\AgdaFunction{controlled} &:& (a \leftrightarrow a) \rightarrow 
                              \bt \otimes a \leftrightarrow \bt \otimes a \\
\AgdaFunction{controlled}~\AgdaFunction{f} &=& 
      \bt \otimes a \\
&& \qquad\byiso{\AgdaFunction{unfoldBool} \otimes \AgdaFunction{id}} \\
&& (\ot \oplus \ot) \otimes a \\
&& \qquad\byiso{\AgdaFunction{dist}} \\
&& (\ot \otimes a) \oplus (\ot \otimes a) \\
&& \qquad\byiso{\AgdaFunction{id} \oplus (\AgdaFunction{id} \otimes \AgdaFunction{f})} \\
&& (\ot \otimes a) \oplus (\ot \otimes a) \\
&& \qquad\byiso{\AgdaFunction{factor}} \\
&& (\ot \oplus \ot) \otimes a \\
&& \qquad\byiso{\AgdaFunction{foldBool} \otimes \AgdaFunction{id}} \\
&& \bt \otimes a \\
\\ 
\AgdaFunction{not} &:& \bt \leftrightarrow \bt \\
\AgdaFunction{not} &=& 
  \AgdaFunction{unfoldBool} \odot \AgdaFunction{swap₊} \odot \AgdaFunction{foldBool} \\
\\ 
\AgdaFunction{cnot} &:& \bt \otimes \bt \leftrightarrow \bt \otimes \bt \\
\AgdaFunction{cnot} &=& \AgdaFunction{controlled}~\AgdaFunction{not} \\
\\ 
\AgdaFunction{toffoli} &:& \bt \otimes (\bt \otimes \bt)
                           \leftrightarrow  \bt \otimes (\bt \otimes \bt) \\
\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot}
\end{array}\]}

@@

\jacques{Refer to some Sabry et al. papers regarding Toffoli gates.}

>>>>

The examples below assume a representation of the type of booleans
$\bt$ as the disjoint union $\ot \oplus \ot$ with the left injection
representing $\mathsf{false}$ and the right injection representing
$\mathsf{true}$. Given an arbitrary reversible function
$\AgdaFunction{f}$ of type $a \leftrightarrow a$, we can build the
reversible function $\AgdaFunction{controlled}~\AgdaFunction{f}$ that
takes a pair of type $\bt \otimes a$ and checks the incoming boolean;
if it is false (i.e., we are in the left injection), the function
behaves like the identity; otherwise the function applies
$\AgdaFunction{f}$ to the second argument. In either case, the
incoming boolean is reconstituted to maintain reversibility. Using the
\AgdaFunction{controlled} building block, we can build a
controlled-not $\AgdaFunction{cnot}$ gate and a
controlled-controlled-not gate, also known as the
\AgdaFunction{toffoli} gate. The latter gate is a universal function
for combinational boolean circuits thus showing the expressiveness of
the language:

\begin{figure}
\begin{center}
\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.9}]
  \draw (1,2) ellipse (0.5cm and 0.5cm);
  \draw[fill] (1,2) circle [radius=0.025];
  \node[below] at (1,2) {*};

  \draw (0,0) ellipse (0.5cm and 1cm);
  \draw[fill] (0,0.5) circle [radius=0.025];
  \node[below] at (0,0.5) {F};
  \draw[fill] (0,-0.5) circle [radius=0.025];
  \node[below] at (0,-0.5) {T};

  \draw     (1,2)    -- (2,2)      ; %% ()
  \draw     (0,0.5)  -- (2,0.5)    ; %% F
  \draw     (0,-0.5) -- (2,-0.5)   ; %% T

  \draw     (2,2)    -- (3,-0.5)   ;
  \draw     (2,0.5)  -- (3,2)      ;
  \draw     (2,-0.5) -- (3,1)      ;

  \draw     (3,2)    -- (3.5,2)    ;
  \draw     (3,1)    -- (3.5,1)    ;
  \draw     (3,-0.5) -- (3.5,-0.5) ;

  \draw     (3.5,2)    -- (4.5,1)    ;
  \draw     (3.5,1)    -- (4.5,2)    ;
  \draw     (3.5,-0.5) -- (4.5,-0.5) ;

  \draw     (4.5,2)    -- (5,2)    ;
  \draw     (4.5,1)    -- (5,1)    ;
  \draw     (4.5,-0.5) -- (5,-0.5) ;

  \draw     (5,2)    -- (6,0.5)  ;
  \draw     (5,1)    -- (6,-0.5) ;
  \draw     (5,-0.5) -- (6,2)    ;

  \draw     (6,2)    -- (7,2)    ;
  \draw     (6,0.5)  -- (8,0.5)  ;
  \draw     (6,-0.5) -- (8,-0.5) ;

  \draw (7,2) ellipse (0.5cm and 0.5cm);
  \draw[fill] (7,2) circle [radius=0.025];
  \node[below] at (7,2) {*};

  \draw (8,0) ellipse (0.5cm and 1cm);
  \draw[fill] (8,0.5) circle [radius=0.025];
  \node[below] at (8,0.5) {F};
  \draw[fill] (8,-0.5) circle [radius=0.025];
  \node[below] at (8,-0.5) {T};

\end{tikzpicture}
\end{center}
\caption{\label{fig:not}Graphical representation of \AgdaFunction{not₃}}
\end{figure}

{\small
\[\def\arraystretch{1.2}\begin{array}{rcl}
\AgdaFunction{controlled} &:& \forall a.~ (a \leftrightarrow a) \rightarrow 
                              (\bt \otimes a \leftrightarrow \bt \otimes a) \\
\AgdaFunction{controlled}~\AgdaFunction{f} &=& 
      \bt \otimes a \\
&& \qquad\byiso{\AgdaFunction{unfoldBool} \otimes \AgdaFunction{id}} \\
&& (\ot \oplus \ot) \otimes a \\
&& \qquad\byiso{\AgdaFunction{distribute}} \\
&& (\ot \otimes a) \oplus (\ot \otimes a) \\
&& \qquad\byiso{\AgdaFunction{id} \oplus (\AgdaFunction{id} \otimes \AgdaFunction{f})} \\
&& (\ot \otimes a) \oplus (\ot \otimes a) \\
&& \qquad\byiso{\AgdaFunction{factor}} \\
&& (\ot \oplus \ot) \otimes a \\
&& \qquad\byiso{\AgdaFunction{foldBool} \otimes \AgdaFunction{id}} \\
&& \bt \otimes a \\
\\ 
\AgdaFunction{not} &:& \bt \leftrightarrow \bt \\
\AgdaFunction{not} &=& 
  \AgdaFunction{unfoldBool} \odot \AgdaFunction{swap₊} \odot \AgdaFunction{foldBool} \\
\\ 
\AgdaFunction{cnot} &:& \bt \otimes \bt \leftrightarrow \bt \otimes \bt \\
\AgdaFunction{cnot} &=& \AgdaFunction{controlled}~\AgdaFunction{not} \\
\\ 
\AgdaFunction{toffoli} &:& \bt \otimes (\bt \otimes \bt)
                           \leftrightarrow  \bt \otimes (\bt \otimes \bt) \\
\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot} \\
\end{array}\]}
%%%

The code for $\AgdaFunction{controlled}~\AgdaFunction{f}$ provides
constructive evidence (i.e., a program, a logic gate, or a hardware
circuit) for an automorphism on $\bt \otimes a$: it can be read
top-down or bottom-up to go back and forth. This perspective naturally
raises the questions of when two programs are ``equivalent.'' Consider
the six programs of type $\bt \leftrightarrow \bt$ below:

{\small 
\[\def\arraystretch{1.2}\begin{array}{rcl}
\AgdaFunction{id₁}~\AgdaFunction{id₂}~\AgdaFunction{id₃}~ 
  \AgdaFunction{not₁}~\AgdaFunction{not₂}~\AgdaFunction{not₃} &:& \bt \leftrightarrow \bt \\
\AgdaFunction{id₁} &=& 
  \AgdaFunction{id} \odot \AgdaFunction{id} \\
\AgdaFunction{id₂} &=& 
  \AgdaFunction{not} \odot \AgdaFunction{id} \odot \AgdaFunction{not} \\
\AgdaFunction{id₃} &=& 
  \AgdaFunction{uniti⋆} \odot \AgdaFunction{swap⋆} \odot 
                        (\AgdaFunction{id} \otimes \AgdaFunction{id}) \odot 
                        \AgdaFunction{swap⋆} \odot 
                        \AgdaFunction{unite⋆} \\
\AgdaFunction{not₁} &=& 
  \AgdaFunction{id} \odot \AgdaFunction{not} \\
\AgdaFunction{not₂} &=& 
  \AgdaFunction{not} \odot \AgdaFunction{not} \odot \AgdaFunction{not} \\
\AgdaFunction{not₃} &=& 
  \AgdaFunction{uniti⋆} \odot \AgdaFunction{swap⋆} \odot 
                        (\AgdaFunction{not} \otimes \AgdaFunction{id}) \odot 
                        \AgdaFunction{swap⋆} \odot 
                        \AgdaFunction{unite⋆} 
\end{array}\]}

\noindent We neither expect, nor desire, that all these six programs
be equivalent. However each group of three programs arguably
represents ``the same'' program. Thinking extensionally, it is indeed
easy to verify this fact but we can do better. We can provide
\emph{evidence} (i.e., a program that manipulates programs) that can
constructively transform one program to an equivalent one. We show
such a level-2 program proving that $\AgdaFunction{not₃}$ is
equivalent to $\AgdaFunction{not}$:

{\small 
\[\def\arraystretch{1.2}\begin{array}{rcl}
\AgdaFunction{notOpt} &:& \AgdaFunction{not₃} \Leftrightarrow \AgdaFunction{not} \\
\AgdaFunction{notOpt} &=& 
  \AgdaFunction{uniti⋆} \odot (\AgdaFunction{swap⋆} \odot 
                        ((\AgdaFunction{not} \otimes \AgdaFunction{id}) \odot 
                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}))) \\
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocLeft}} \\
&& \AgdaFunction{uniti⋆} \odot (\AgdaFunction{swap⋆} \odot 
                        (\AgdaFunction{not} \otimes \AgdaFunction{id})) \odot 
                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}) \\
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{swapLeft} 
                                  \boxdot \AgdaFunction{id})} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot \AgdaFunction{swap⋆}) \odot 
                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}) \\
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocRight}} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot (\AgdaFunction{swap⋆} \odot 
                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}))) \\
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
                                  \boxdot \AgdaFunction{assocLeft})} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot ((\AgdaFunction{swap⋆} \odot 
                      \AgdaFunction{swap⋆}) \odot \AgdaFunction{unite⋆})) \\
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
                                  \boxdot (\AgdaFunction{leftInv} \boxdot \AgdaFunction{id}))} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot (\AgdaFunction{id} \odot \AgdaFunction{unite⋆})) \\
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
                                  \boxdot \AgdaFunction{idLeft})} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot \AgdaFunction{unite⋆}) \\
&& \qquad\byisotwo{\AgdaFunction{assocLeft}} \\
&& (\AgdaFunction{uniti⋆} \odot (\AgdaFunction{id} \otimes \AgdaFunction{not})) 
                      \odot \AgdaFunction{unite⋆} \\
&& \qquad\byisotwo{\AgdaFunction{unitiLeft} \boxdot \AgdaFunction{id}} \\
&& (\AgdaFunction{not} \otimes \AgdaFunction{uniti⋆}) \odot \AgdaFunction{unite⋆} \\    
&& \qquad\byisotwo{\AgdaFunction{assocRight}} \\
&& \AgdaFunction{not} \otimes (\AgdaFunction{uniti⋆} \odot \AgdaFunction{unite⋆}) \\    
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{leftInv}} \\
&& \AgdaFunction{not} \otimes \AgdaFunction{id} \\
&& \qquad\byisotwo{\AgdaFunction{idRight}} \\
&& \AgdaFunction{not} 
\end{array}\]}

%% \jacques{Example 1: using the combinators from pi-dual, some an equational
%% derivation in type theory (which?), in equational-derivation style.  Point out
%% that reading ``down'' the justifications is a $\Pi$ program which is also a
%% constructive proof of the equivalence of the types.  As it is reversible, we
%% can also read the reverse program by reading ``up''.}

@@

%% \jacques{Refer to some Sabry et al. papers regarding Toffoli gates.}
